Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-06 manuscript/manuscript.tex
Finish non-recombining sites.

[fpmi vs tajima's D]

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/figure2.pdf}
\caption{[CAPTION] \label{fig:pmi_vs_tajimasd}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/trees.png}
\caption{Genealogies with a sample size of $n=4$. Opportunities for singleton/tripleton mutations are in orange. Opportunities for doubletons are in blue. (Note, there is a third multiple-merger topology, not shown.) \label{fig:trees}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/figure3.pdf}
\caption{[CAPTION] \label{fig:sdpmi_vs_sdratio}}
\end{figure}

[Singleton/doubleton fpmi versus singleton/doubleton ratio]

>>>>

Accordingly, we computed the coarse-grained $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
For the same parameters, we also calculated a normalized Tajima's D statistic:
$(\hat{\theta}_W - \pi) / \pi$, where $\hat{\theta}_W$ is Watterson's estimator of the coalescent scaled mutation rate \autocite{Watterson19??}.
The latter is a measure of the distortions in the SFS induced by population growth or multiple mergers and equals zero for the constant-$N$ Kingman coalescent.
\Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
On the other hand, beta coalescents that generate similar distortions in the SFS, generate larger changes in $\hilopmi$, including positive values.
Thus, the 2-SFS in general, and $\hilopmi$ in particular, are capable of capturing the effects of multiple mergers beyond the distortions in branch lengths.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/figure2.pdf}
\caption{Pointwise mutual information between singletons and non-singletons for Kingman coalescents with exponential growth and beta coalescents. The exponential growth rate, $g$, ranges from 0.25 to 8.0 in coalescent time units. The beta coalescent parameter, $\alpha$, ranges from 1.75 (nearly Kingman) to 1.25 (nearly Bolthausen-Sznitman). \label{fig:fpmi_vs_tajimasd}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/trees.png}
\caption{Genealogies with a sample size of $n=4$. Opportunities for singleton/tripleton mutations are in orange. Opportunities for doubletons are in blue. (Note: there is a third possible multiple-merger topology, not shown.) \label{fig:trees}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/figure3.pdf}
\caption{Distortions in $\fpmi$ vs. distortions in the SFS for $n=4$. Parameters for beta coalescents and exponential growth are as in \fig{fig:fpmi_vs_tajimasd}. For the piecewise-constant--$N$ model, the fold-change in $N$ varies from 2 to $2^4$ and the time of this change varies from $2^-4$ to 2 in coalescent time units. \label{fig:sdpmi_vs_sdratio}}
\end{figure}

As argued above, $\E{\eta_{12}}$ is also distorted by changes in the coalescent time distribution induced by population growth.
\Fig{fig:sdpmi_vs_sdratio} demonstrates that $\fpmi$ accounts for this fact by normalizing by the SFS.
\Fig{fig:sdpmi_vs_sdratio} plots $\fpmi_{12}$ against the ratio of singletons to doubletons $\eta_1 / \eta_2$ for the beta coalescent and two models of population growth: exponential growth and a piecewise-constant model with two epochs.
In the latter model, we vary both the fold-change in $N$ and the time of the change.
Like Tajima's D in the previous section, the singleton/doubleton ratio measures distortion the SFS relative to the constant-$N$ Kingman coalescent.
(In fact, with $n=4$, this ratio captures \emph{all} of the distortion in the SFS.)
As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
Moreover, the two different models of growth collapse onto one another.
This suggests that for the Kingman coalescent with growth, $\fpmi_{12} \approx f(\eta_1/\eta_2)$ for some function $f$.
